---
title: "R Notebook"
output: html_notebook
---

##Analytical Query I: Top five journals with the most articles published in them for the time period. Think about how you would best present this information so it is easily usable by an analyst or manager.

```{r}
##Connect to Database
library(DBI)
library(RMySQL)
library(tinytex)
options("mysql.local_infile" = TRUE)
dbcon<-dbConnect(MySQL(),host="localhost",dbname="journal_star_db",user="root",
                 password="Asm@rino123",port=3306)
dbcon
```
### So for obtaining Top five journals with the most articles published in them for the given time period we have analyzed the fact table f_journal and dimension table d_time and by having the partitions of journals with respect to time period that is with {year,month}. And within each of the partitions picking up *top 5 journal which are having highest number of articles*

### Here are the first 10 records from the resultant dataframe:

```{r}
top5_journals_df = dbGetQuery(dbcon, "select year,month,
                                journalId,title,num_of_articles
                                from 
                               (SELECT ROW_NUMBER() OVER (PARTITION BY j.t_id
                                     order by j.num_of_articles desc) as rn,
                                     j.*
                            FROM f_journal j)tmp 
                            join d_time t
                            on tmp.t_id=t.t_id
                            order by year,month,num_of_articles desc
                            ")

head(top5_journals_df,10)
```

### Lets consider the time period of **1977** and **January** month to see the top 5 journals having highest number of articles 

```{r}
top5_journals_df = dbGetQuery(dbcon, "select year,month,
                                journalId,title,num_of_articles
                                from 
                               (SELECT ROW_NUMBER() OVER (PARTITION BY j.t_id
                                     order by j.num_of_articles desc) as rn,
                                     j.*
                            FROM f_journal j)tmp 
                            join d_time t
                            on tmp.t_id=t.t_id
                            where tmp.rn<6 and year=1977 
                            and month in ('jan')
                            order by year,month,num_of_articles desc
                            ")

top5_journals_df
```
```{r}
#install.packages("ggplot2")
library("ggplot2")

ggplot(top5_journals_df, aes(x=title, y=num_of_articles,
                                 fill=num_of_articles))+
  geom_bar(stat='identity')+
  theme(axis.text.x = element_text(angle = 60))
```
### From the above plot we can see that the journal **Journal of applied physiology: respiratory, environmental and exercise physiology** has significantly published many articles during 1977 in the month of January

##Analytical Query II: Number of articles per journal per year broken down by quarter. Again, think of a good way to show this. Is a table better than a visualization or is it better to use a visualization.

### To obtain the Number of articles per journal per year broken down by quarter we analyzed the fact table f_journal and dimension table d_time tables, in order to get the number of articles per journal per year and per quarter wise - we grouped the data based on `year`, `quarter` and `journal`, and then aggregated to obtain the **number of articles** with in that group

### Here are the first 5 records from the resultant dataframe:

```{r}
num_artcls_df = dbGetQuery(dbcon, "select 
                                journalId,title,year,quarter,
                                sum(num_of_articles) as num_of_articles
                                from f_journal j
                                join d_time t
                                on j.t_id=t.t_id
                                group by t.year,t.quarter,j.journalId,j.title
                                order by year,journalId
                            ")

head(num_artcls_df,5)
```

### Lets consider to find the *highest* number of articles publishing journals per year and per quarter

```{r}
num_artcls_df = dbGetQuery(dbcon, "select 
                                journalId,title,year,quarter,
                                sum(num_of_articles) as num_of_articles
                                from f_journal j
                                join d_time t
                                on j.t_id=t.t_id
                                group by t.year,t.quarter,j.journalId,j.title
                                order by num_of_articles desc limit 7
                            ")

num_artcls_df
```
```{r}
#install.packages("ggplot2")
library("ggplot2")

ggplot(num_artcls_df, aes(x=title, y=num_of_articles,
                                 fill=num_of_articles))+
  geom_bar(stat='identity')+
  theme(axis.text.x = element_text(angle = 60))
```

### So from the above graph we can see that the journal *Biochimica et biophysica acta* and *The Journal of pharmacy and pharmacology* have published more than **750** articles and have been the top journal which published higheer number of articles.
