---
title: "Practicum I CS5200"
author: "Samuel Ghebreyesus"
date: "Spring 2023"
output:
  pdf_document: default
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

## Connect to Database 
```{r connectDB, eval=TRUE, warning=FALSE}
library(RMySQL)
conn = dbConnect(RMySQL::MySQL(), host='localhost',
                            port=3306, user='root', 
                 password='Asm@rino123')
dbSendQuery(conn, "CREATE DATABASE birdstrikedb")
dbSendQuery(conn, "USE birdstrikedb")
dbSendQuery(conn, "SET GLOBAL local_infile=1")
conn = dbConnect(RMySQL::MySQL(), dbname='birdstrikedb', host='localhost',
                            port=3306, 
                 user='root', password='Asm@rino123')
```


## Create Database
```{r createDB, eval=TRUE, warning=FALSE}
sql <- "DROP TABLE IF EXISTS airports, conditions, airlines, incidents"
dbSendStatement(conn, sql)

airquery <- "CREATE TABLE airports(
  aid INT NOT NULL AUTO_INCREMENT,
  airportName TEXT NOT NULL,
  airportCode TEXT,
  state TEXT NOT NULL,
  PRIMARY KEY (aid)
)"

airresults <- dbSendQuery(conn, airquery)
dbClearResult(airresults)

condquery <- "CREATE TABLE conditions(
  cid INTEGER NOT NULL,
  `condition` TEXT NOT NULL,
  explanation TEXT,
  PRIMARY KEY (cid)
)"

condresults <- dbSendQuery(conn, condquery)
dbClearResult(condresults)

airlinequery <- "CREATE TABLE airlines(
  eid INT NOT NULL AUTO_INCREMENT,
  airlineName TEXT NOT NULL,
  airlineCode TEXT,
  flag TEXT,
  PRIMARY KEY (eid)
)"

airlineresults <- dbSendQuery(conn, airlinequery)
dbClearResult(airlineresults)

incquery <- "CREATE TABLE incidents(
  rid INTEGER NOT NULL,
  `dep.date` DATE,
  origin INTEGER,
  airline INTEGER,
  aircraft TEXT,
  `flight.phase` ENUM('takeoff', 'landing', 'inflight', 'unknown'),
  altitude INTEGER,
  CHECK (altitude >= 0),
  conditions INTEGER,
  warned TEXT,
  PRIMARY KEY (rid),
  FOREIGN KEY (origin) REFERENCES airports (aid),
  FOREIGN KEY (conditions) REFERENCES conditions (cid),
  FOREIGN KEY (airline) REFERENCES airlines (eid)
)"

incresults <- dbSendQuery(conn, incquery)
dbClearResult(incresults)
```


## Tests table definition; attempt to insert a negative number into altitude column, which is illegal and should return an error 
```{r testtableDef, eval=FALSE}
sql <- "insert into incidents (rid, altitude) values (9999999, -1)"
dbSendQuery(conn, sql)
```

## Tests table definition; attempt to insert a value into flight phase column that isn't takeoff, landing, inflight or unknown, should return error
```{r testtableDefTwo, eval=FALSE}
sql <- "insert into incidents(rid, `flight.phase`) 
values (10000000, 'driving');"
dbSendQuery(conn, sql)
```


## Load birdstrikes csv data file into a dataframe called bds.raw. Printed out first 5 rows just to show that it worked. 
```{r loadData, eval=TRUE, warning=FALSE}
fn = "BirdStrikesData-V2.csv"
bds.raw <- read.csv(file = fn,
                   header = T,
                   stringsAsFactors = F)
head(bds.raw, 5)
```


## Populate the tables with the data from the appropriate dataframe columns. I went through each row and where there was no airport or airline, I put in the value 'Unknown'. Used sqldf to get the necessary subset of the csv file and bulk loaded the subset into the tables. 
```{r populateAirports, eval=TRUE, warning=FALSE}
options(sqldf.driver = 'SQLite')
df.airports <- sqldf::sqldf("select 1 as aid, 
                            airport as 'airportName', 
                            '' as airportCode, 
                            origin as 'state' from `bds.raw` 
                            group by airport, state")
numairports <- nrow(df.airports)
df.airports[,1] <- seq(1, numairports)
for (r in 1:numairports) {
  if (df.airports$airportName[r] == '') {
    df.airports$airportName[r] = 'Unknown'
  }
}
dbWriteTable(conn, "airports", df.airports, overwrite = F, append = T, 
             row.names = FALSE)
```

```{r populateAirlines, eval=TRUE, warning=FALSE}
df.airlines <- sqldf::sqldf("select 1 as eid, 
                            airline as 'airlineName', 
                            '' as airlineCode, 
                            '' as 'flag' from `bds.raw` 
                            group by airline")
numairlines <- nrow(df.airlines)
df.airlines[,1] <- seq(1, numairlines)
for (r in 1:numairlines) {
  if (df.airlines$airlineName[r] == '') {
    df.airlines$airlineName[r] = 'Unknown'
  }
}
dbWriteTable(conn, "airlines", df.airlines, overwrite = F, append = T, 
             row.names = FALSE)
```

```{r populateConditions, eval=TRUE, warning=FALSE}
df.conditions <- sqldf::sqldf("select 1 as cid, sky_conditions as 'condition', 
                              '' as explanation from `bds.raw` group by sky_conditions")
numconditions <- nrow(df.conditions)
df.conditions[,1] <- seq(1, numconditions)
dbWriteTable(conn, "conditions", df.conditions, overwrite = F, append = T, 
             row.names = FALSE)
```

## This code chunk was used to populate the incidents table. I had to find the aid, eid and cid values that matched the airport/airline names and conditions to satisfy the foreign key constraint. I also had to map the flight phase values to their proper value in the value set. Based on research, I mapped approach, descent and landing roll to landing. I also mapped Climb to inflight, Take-off run to takeoff and everything else(taxi, parked, blank) to unknown 
```{r populateIncidents, eval=TRUE, warning=FALSE}
df.incidents <- sqldf::sqldf("select * from `bds.raw`")
numincidents <- nrow(df.incidents)
for (r in 1:numincidents) {
  if (df.incidents$airport[r] == '') {
    df.incidents$airport[r] = 'Unknown'
  }
  if (df.incidents$airline[r] == '') {
    df.incidents$airline[r] = 'Unknown'
  }
    a <- df.airports$aid[which(df.incidents$airport[r] == 
                                 df.airports$airportName 
                               & df.incidents$origin[r] == df.airports$state)]
    b <- df.airlines$eid[which(df.incidents$airline[r] == 
                                 df.airlines$airlineName)]
    c <- df.conditions$cid[which(df.incidents$sky_conditions[r] == df.conditions$condition)]
    df.incidents$airport[r] = a
    df.incidents$airline[r] = b
    df.incidents$sky_conditions[r] = c
    if (df.incidents$flight_phase[r] == 'Approach' 
        || df.incidents$flight_phase[r] == 'Descent' 
        || df.incidents$flight_phase[r] == 'Landing Roll') {
      df.incidents$flight_phase[r] = 'landing'
    } else if (df.incidents$flight_phase[r] == 'Climb') {
      df.incidents$flight_phase[r] = 'inflight' 
    } else if (df.incidents$flight_phase[r] == 'Take-off run') {
      df.incidents$flight_phase[r] = 'takeoff'
    } else {
      df.incidents$flight_phase[r] = 'unknown'
    }
}
df.incidents2 <- sqldf::sqldf("select rid, 
                              flight_date as `dep.date`, 
                              airport as 'origin',
                              airline, aircraft, 
                              flight_phase as `flight.phase`, 
                              altitude_ft as 'altitude', 
                              sky_conditions as 'conditions', 
                              pilot_warned_flag as 'warned' 
                              from `df.incidents`")
df.incidents2$dep.date = substr(df.incidents2$dep.date, 1, 
                            nchar(df.incidents2$dep.date)-5)
df.incidents2$dep.date <- as.Date(df.incidents2$dep.date, format = "%m/%d/%Y")
df.incidents2$altitude <- gsub(',', '', df.incidents$altitude)
dbWriteTable(conn, "incidents", df.incidents2, overwrite = F, append=T, 
             row.names = FALSE)
```


## Display the first five rows of each table to show that the data loading worked
```{r dispTable, eval=TRUE, warning=FALSE}
airports <- dbGetQuery(conn, "select * from airports")
head(airports, 5)

airlines <- dbGetQuery(conn, "select * from airlines")
head(airlines, 5)

conditions <- dbGetQuery(conn, "select * from conditions")
head(conditions, 5)

incidents <- dbGetQuery(conn, "select * from incidents")
incidents$dep.date = as.Date(incidents$dep.date)
head(incidents, 5)
```


## SQL query to find the 10 states with the greatest number of incidents
```{sql greatestIncidentStates, connection=conn}
select state, count(rid) as 'numIncidents' from (
  select incidents.rid, airports.state 
  from incidents
  inner join airports on incidents.origin = airports.aid) as rid_state
  group by rid_state.state
  order by numIncidents desc limit 10
```


## SQL query to find the airlines that had an above average number bird strike incidents
```{sql aboveAvgAirlines, connection=conn}
select * from (
select airlines.airlineName, count(incidents.rid) as 'numIncidents'
from airlines
inner join incidents on incidents.airline = airlines.eid
group by airlines.eid) as incident_airline_table
where numIncidents > (select count(rid) / count(distinct airline) from incidents)
order by numIncidents asc
```


## SQL query to find the number of bird strike incidents by month and by flight phase, across all years and including all airlines and flights. I excluded from the data frame the rows that had unknown month as that would not be useful data for the analytics we are trying to run. 
```{r groupbyMonthandPhases, eval=TRUE, warning=FALSE}
sql <- "select MONTH(`dep.date`) as 'Month', `flight.phase`, count(rid) as 'numIncidents' 
from incidents
where MONTH(`dep.date`) IS NOT NULL
group by Month, `flight.phase`
order by Month, `flight.phase`"
month_phase_df_display <- dbGetQuery(conn, paste(sql, "limit 6"))
month_phase_df_full <- dbGetQuery(conn, sql)
month_phase_df_display
```


## Build a scatter plot that plots month along the x-axis versus number of incidents across all airlines and flight phases on the y-axis 
```{r buildScatterPlot, eval=TRUE, warning=FALSE}
sum_df <- aggregate(month_phase_df_full$numIncidents, list(month_phase_df_full$Month), 
                    FUN=sum)
plot(sum_df$Group.1, sum_df$x, xlab="Month", 
    ylab="Total # of Birdstrike Incidents", 
    main="Total Number of Birdstrike Incidents By Month", 
    pch=19, xaxt='n')
axis(side=1, at=seq(1, 12, by=1))
```


## Stored procedure in MySQL that adds a new incident to the database. It takes in as parameters airport name, state, airline name, rid, date, aircraft, flight phase, condition and warned flag (y/n). If the airport or airline does not already exist in their respective tables then they are newly added to the airports/airlines table.
```{sql createProc, connection = conn}
CREATE PROCEDURE INSERTNEWINCIDENT(
    IN newairportName TEXT,
    IN newstate TEXT,
    IN newairlineName TEXT, 
    IN newrid INT,
    IN newdate DATE,
    IN newaircraft TEXT,
    IN newflightphase TEXT,
    IN newaltitude INT,
    IN newconditions INT,
    IN newWarned TEXT
)
BEGIN
    IF NOT EXISTS (SELECT * FROM AIRPORTS 
    WHERE airportName = newairportName and 
    state = newstate) THEN
        INSERT INTO airports (airportName, state)
        VALUES (newairportName, newstate);
    END IF;
    IF NOT EXISTS (SELECT * FROM AIRLINES WHERE airlineName = newairlineName) 
    THEN
        INSERT INTO AIRLINES (airlineName) 
        VALUES (newairlineName);
    END IF;
    INSERT INTO INCIDENTS(rid, `dep.date`, 
    origin, airline, aircraft, `flight.phase`, 
    altitude, conditions, warned) 
    VALUES(newrid, newdate, (select aid from airports
where airportName = newairportName and state = newstate), (select eid from 
airlines where airlineName = newairlineName), 
newaircraft, newflightphase, newaltitude, 
newconditions, newWarned);
END; 
```


## Calling my stored procedure to insert these 5 incidents. The first three incidents contain new airports, with two of them also including new airlines. The last two incidents contain airlines/airports that already exist. 
```{sql insertIncident, connection = conn}
CALL INSERTNEWINCIDENT("Asmara International", "Africa", "Africa Airways", 
9999997, "2010-11-19", "Airplane", "unknown", 200, 2, "Y");
```

```{sql insertIncidentTwo, connection=conn}
CALL INSERTNEWINCIDENT("Addis Airport", "Africa", "Ethiopia Airlines", 
9999998, "2016-11-19", "Airplane", "unknown", 1200, 3, "N");
```

```{sql insertIncidentThree, connection=conn}
CALL INSERTNEWINCIDENT("Unknown", "Michigan", "ACM AVIATION", 9999999, 
"2014-11-16", "Airplane", "landing", 1700, 1, "N");
```

```{sql insertIncidentFour, connection=conn}
CALL INSERTNEWINCIDENT("ANN ARBOR MUNICIPAL", "Michigan", "ACM AVIATION", 
10000000, "2014-11-16", "Airplane", "landing", 1500, 3, "Y");
```

```{sql insertIncidentFive, connection=conn}
CALL INSERTNEWINCIDENT("Unknown", "Maryland", "American Airlines", 
10000001, "2014-11-16", "Airplane", "landing", 1500, 3, "N");
```


## Testing that the 3 new airports were added to the airports table and thus that the stored procedure worked 
```{r testAirportsAdded, eval=TRUE, warning=FALSE}
sql <- "select * from airports
where airportName = 'Asmara International' 
or airportName = 'Addis Airport' 
or (airportName = 'Unknown' and state ='Michigan');"
dbGetQuery(conn, sql)
```

## Testing that the 2 new airlines were added to the airlines table and thus that the stored procedure worked 
```{r testAirlinesAdded, eval=TRUE, warning=FALSE}
sql <- "select * from airlines
where airlineName = 'Africa Airways' or airlineName = 'Ethiopia Airlines'"
dbGetQuery(conn, sql)
```

## Testing that the 5 incidents were added to the incidents table and thus that the stored procedure worked 
```{r testIncidentsAdded, eval=TRUE, warning=FALSE}
sql <- "select * from incidents
where rid >= 9999997 and rid <= 10000001;"
dbGetQuery(conn, sql)
```

## Test that only the three new airports were inserted; original airlines table had 1141 rows
```{r testcountAirports, eval=TRUE, warning=FALSE}
sql <- "select count(*) from airports;"
dbGetQuery(conn, sql)
```

## Test that only the two new airlines were inserted; original airports table had 293 rows
```{r testcountAirlines, eval=TRUE, warning=FALSE}
sql <- "select count(*) from airlines;"
dbGetQuery(conn, sql)
```

## Test that 5 incidents were inserted; original incidents table had 25558 rows
```{r testCountIncidents, eval=TRUE, warning=FALSE}
sql <- "select count(*) from incidents"
dbGetQuery(conn, sql)
```


